\documentclass[]{article}
\usepackage{amssymb,amsmath,mathrsfs,bm}
\usepackage{hyperref}
\usepackage{xcolor}

%opening
\title{Notes on Minimax filtering}
\author{Albert Akhriev}

\newcommand{\TODO}[1]{({\bf{\color{red}{TODO:}}}\,\,\,{\footnotesize\it{#1}})}

\begin{document}
\maketitle

%\begin{abstract}
%\end{abstract}

\section{Minimax filter basics}

To be done ...

%Source: \cite{wiki1}, \cite{Welch06}.
%\pagebreak

\subsection{Minimax filter model}

In what follows, we consider a simplified version of the discrete time minimax filter \cite{Zhuk09}.

\noindent\textbf{Process model}:
\begin{equation}
x_{k+1} = C_k x_k + f_{k+1}.
\end{equation}
\noindent\textbf{Measurement model}:
\begin{equation}
y_k = H_k x_k + g_k.
\end{equation}
\begin{description}
\item[$x_k$]: state vector of the system;
\item[$C_k$]: state transition model applied to the previous state $x_{k-1}$;
\item[$H_k$]: observation model that maps the state to observations;
\item[$f_k$]: process noise with zero mean and covariance $S_k$. \TODO{true?}
\item[$g_k$]: measurement noise with zero mean and covariance $R_k$; \TODO{true?}
\end{description}

\noindent Noises: (1) independent of each other; (2) not necessary Gaussian. \TODO{true?}

\subsection{Minimax filter algorithm}

\noindent\textbf{Initialization}:
\begin{eqnarray}
r_0      &=& H_0^T R_0 \, y_0, \\
P_0      &=& H_0^T R_0 \, H_0 + S_0, \\
\alpha_0 &=& y_0^T R_0 \, y_0, \\
B_0      &=& P_0 + C_0^T S_0 \, C_0,  \mbox{\TODO{true?}}
%\beta_0  &=& 1 - \alpha_0 + x_0^T P_0 \, x_0.
\end{eqnarray}

\noindent\textbf{Progression}: \TODO{$C_{k-1} B_{k-1}^{+} C_k^T$  or  $C_{k-1}^T B_{k-1}^{+} C_k$ ???}
\begin{eqnarray}
P_k      &=& H_k^T R_k \, H_k + S_{k-1} - S_{k-1}^T C_{k-1} B_{k-1}^{+} C_k^T S_{k-1},   \\
B_k      &=& P_k + C_k^T S_k \, C_k, \\
\alpha_k &=& \alpha_{k-1} + y_k^T R_k \, y_k - r_{k-1}^T B_{k-1}^{+} \, r_{k-1}, \\
r_k      &=& S_{k-1} C_{k-1} B_{k-1}^{+} r_{k-1} + H_k^T R_k \, y_k, \\
x_k      &=& P_k^{+} \, r_k, \\
\beta_k  &=& 1 - \alpha_k + x_k^T P_k \, x_k.
\end{eqnarray}

\noindent\textbf{Post-condition}: $\beta_k \ge 0$, otherwise we claim filter failure. \TODO{true?}

\begin{thebibliography}{1}
\bibitem{Zhuk09} Zhuk~S.M., ``Minimax state estimation for linear discrete-time differential-algebraic equations'', Automatica J. IFAC, 46 (11), pp.1785-1789, 2010.
\end{thebibliography}

\end{document}
